---
title: "Choosing the grain a sensor record is read at"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the grain a sensor record is read at}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
set.seed(1)
library(timegrain)
```

A logger records every hour for years. Before any model sees it, the record is reduced: to monthly
means, to growing-degree-days, to whatever the analyst settled on once. `timegrain` makes that
reduction an argument, fits at each setting of it, and shows where predictive skill saturates.

This vignette runs the whole path on a small simulated record, from readings to a curve that says
how much of the record the models needed.

## A record to read

Sixty plots, four months of hourly readings. Each plot has a level of its own, and that level is
buried in hour-to-hour noise an order of magnitude larger.

```{r data}
hours <- seq(as.POSIXct("2021-09-01", tz = "UTC"), by = "hour", length.out = 24 * 120)
plots <- sprintf("p%03d", 1:60)
warmth <- rnorm(60)

readings <- data.frame(
  plot = rep(plots, each = length(hours)),
  t = rep(hours, times = 60),
  temp = as.numeric(vapply(warmth, function(w) {
    1.5 * w + 6 * sin(seq_along(hours) / (24 * 40)) + rnorm(length(hours), sd = 12)
  }, numeric(length(hours))))
)

sign <- rep(c(1, -1), length.out = 6)
y <- matrix(rbinom(60 * 6, 1, plogis(3 * as.numeric(outer(warmth, sign)))), ncol = 6,
            dimnames = list(plots, paste0("sp", 1:6)))
```

## The representation

`window_matrix()` bins the readings by the calendar and summarises every bin.

```{r representation}
x <- window_matrix(readings, plot, t, temp, window = "week",
                   stats = c("cold_day", "mean", "warm_day"))
x
```

The result is a `[plot, bin, channel]` array, and it carries the binning that produced it.

```{r bins}
attr(x, "bin_n")[1, 1:4]
dimnames(x)[[3]]
```

Bins follow the calendar. A month is 28, 30 or 31 days, and a week starts on a Monday, so a bin is
a real month or a real week and not a drifting block of 730 or 168 hours.

```{r calendar}
attr(window_matrix(readings, plot, t, temp, window = "month"), "bin_n")[1, ]
```

### An extreme day is not an extreme reading

`min` and `max` take the coldest and warmest single reading of a bin. `cold_day` and `warm_day`
reduce each day to its own mean first and then take the extreme over days. `mean_daily_min` and
`mean_daily_max` take the mean of the daily extremes, which is the exposure a typical day of the
bin brought. One hour at -50 sets `min` to -50 outright; it reaches the day-level statistics only
through its twenty-fourth of that day's mean.

```{r statistics}
week <- window_matrix(readings, plot, t, temp, window = "week",
                      stats = c("min", "mean_daily_min", "cold_day", "mean",
                                "warm_day", "mean_daily_max", "max"))
round(week[1, 1, ], 2)
```

## The split, and the cells a score is defined on

One fold map, built once, is read by everything that scores, so every model is fitted and scored on
identical splits.

```{r folds}
folds <- fold_map(y, v = 5, seed = 1)
folds
```

A per-species score needs both classes among the held-out plots, and a per-species model needs both
classes among the plots it was fitted on. The mask says which cells those are, from the response
and the fold map alone.

```{r cells}
scorable_cells(y, folds)
```

Because it involves no model, every arm below is restricted to the same cells: their means share
one denominator and every paired difference runs on matched cells.

## The ladder

`window_matrix()` given several windows returns one representation per window, which is what
`window_ladder()` fits across.

```{r ladder}
set <- window_matrix(readings, plot, t, temp, window = c("day", "week", "month"))
lad <- window_ladder(set, y, glmnet_learner(), folds = folds, verbose = FALSE)
summary(lad)
```

```{r plot, fig.alt = "Mean true skill statistic against aggregation window, with an interval across species."}
plot(lad)
```

Skill rises as the record is smoothed, and it is still rising at the coarsest window here: the
signal was planted at a scale wider than a month and the daily reading is still buried in the
hourly noise. The paired contrast is what a claim about that rests on. The difference is taken
inside each cell both arms scored, averaged within a species, and summarised across the species,
which are the independent replicates. Six species is few enough that the interval spans zero even
where the means are ordered.

```{r contrast}
paired_contrast(lad, "month|glmnet", "day|glmnet")
```

Where the whole curve is the question rather than one pair of it, `window_contrasts()` fits a mixed
model on the per-cell scores and compares every window against the best one, correcting for the
comparisons made and no others.

```{r window-contrasts, eval = requireNamespace("emmeans", quietly = TRUE) && requireNamespace("lmerTest", quietly = TRUE)}
window_contrasts(lad)
```

## Reading a level honestly

The true skill statistic is read at the threshold that maximises it, chosen on the same plots the
score is then read on. That inflates the level, and by more the fewer presences a cell holds.
`tss_inflation()` measures the inflation for the presence counts of the design in hand.

```{r inflation}
tss_inflation(y, folds, skill = c(0.6, 0.9), replicates = 100)
```

The inflation is common to every arm scored the same way, so it cancels in the paired differences
above. It does not cancel in a level, so a level is an upper bound on the skill a population has.

## Other learners

The learners that ship are the two aggregate-feature arms, `glmnet_learner()` and
`stepwise_learner()`, and three sequence encoders that need `torch`: `mlp_learner()`,
`cnn_learner()` and `rescnn_learner()`. All three encoders are joint multi-label models, so every
species is predicted together from a shared embedding, which is what makes the rarer ones learnable
at these sample sizes.

An encoder that ends in global pooling discards when a thermal event happened, so the position of a
bin in the year has to be given to it as input.

```{r channels}
week_mean <- window_matrix(readings, plot, t, temp, window = "week")
dimnames(bind_channels(week_mean, calendar_channels(week_mean)))[[3]]
```

`ensemble_learner()` averages members' predicted probabilities before the threshold is chosen, so
the set is scored as one model rather than as a vote between decisions. Members differing in
architecture, in width, or in the grain they read are what it is for.

```{r ensemble}
both <- ensemble_learner(list(ridge = glmnet_learner(alpha = 0), lasso = glmnet_learner(alpha = 1)))
summary(window_ladder(set["week"], y, both, folds = folds, verbose = FALSE))
```

A learner of your own is a fit and a predict pair, and nothing else. It goes through the same
folds, the same cells and the same scoring as the ones that ship.

```{r custom}
flat <- function(x) matrix(as.numeric(x), nrow = dim(x)[1])

nearest_neighbour <- learner(
  "1nn",
  fit = function(x, y, ...) list(x = flat(x), y = y),
  predict = function(model, x) {
    d <- as.matrix(dist(rbind(flat(x), model$x)))[seq_len(dim(x)[1]), -seq_len(dim(x)[1])]
    model$y[apply(d, 1, which.min), , drop = FALSE]
  }
)

summary(window_ladder(set, y, list(glmnet = glmnet_learner(), nn = nearest_neighbour),
                      folds = folds, verbose = FALSE))
```

## What was read

With the fits kept, `bin_occlusion()` holds each bin of the record back in turn, rescores the
held-out plots, and records the fall in score as that bin's weight. Nothing is refitted.

```{r occlusion}
kept <- window_ladder(set["month"], y, glmnet_learner(), folds = folds,
                      keep_fits = TRUE, verbose = FALSE)
weight <- bin_occlusion(kept, set, y, "month|glmnet", permutations = 5)
head(aggregate(weight ~ part, weight, mean), 4)
```

Holding a channel back instead asks what each statistic of a window carries, which is the question
behind keeping a bin's extremes at all.
